---
title: "Assignment1-DAV"
author: "EMMANUEL GODWIN BASSEY || S/N: 3092017"
date: "2025-03-03"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# Load required libraries
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(corrplot)
library(stringr)
library(psych)
library(mice)
```

## 1. Load the Data

```{r load-bike_data}
# Parse CSV into bike_data
bike_data <- read.csv("Bike Buyers Assignment 1.csv", stringsAsFactors = FALSE)
str(bike_data)
summary(bike_data)
head(bike_data)
```

## 2. Data Cleaning
```{r clean-bike_data}
# Checking for duplicate IDs
duplicate_count <- sum(duplicated(bike_data$ID))
cat("Duplicate IDs:", duplicate_count, "\n\n")

# Expected values for categorical variables
expected_marital <- c("Married", "Single")
expected_gender <- c("Male", "Female")
expected_education <- c("Bachelors", "Partial College", "High School", "Graduate Degree", "Partial High School")
expected_home_owner <- c("Yes", "No")
expected_commute <- c("0-1 Miles", "1-2 Miles", "2-5 Miles", "5-10 Miles", "10+ Miles")
expected_region <- c("Europe", "North America", "Pacific")
expected_purchase <- c("Yes", "No")

# Function to check incorrect values
check_invalid_values <- function(column, expected) {
  invalid_values <- setdiff(unique(bike_data[[column]]), expected)
  cat("Incorrect values in", column, ":", if(length(invalid_values) == 0) "None" else invalid_values, "\n")
}

# Run checks
check_invalid_values("Marital.Status", expected_marital)
check_invalid_values("Gender", expected_gender)
check_invalid_values("Education", expected_education)
check_invalid_values("Home.Owner", expected_home_owner)
check_invalid_values("Commute.Distance", expected_commute)
check_invalid_values("Region", expected_region)
check_invalid_values("Purchased.Bike", expected_purchase)

# Check numeric columns for negative values
for (col in c("Income", "Children", "Cars", "Age")) {
  invalid_values <- bike_data[[col]][bike_data[[col]] < 0]
  cat("Invalid values in", col, ":", if(length(invalid_values) == 0) "None" else invalid_values, "\n")
}



```

## 3. Identify Missing Values

```{r check-missing}
# Check for missing values in each column, treating empty strings as NA
missing_bike_data <- sapply(bike_data, function(x) sum(is.na(x) | x == ""))
print(missing_bike_data)

# Display total missing values across all columns
total_missing <- sum(missing_bike_data)
cat("Total missing values in bike_data:", total_missing, "\n\n")
print(as.data.frame(missing_bike_data))
```

## 4. Impute Missing Values

```{r impute-missing-values, message=FALSE, warning=FALSE}
# Convert categorical variables to factors
categorical_vars <- c("Marital.Status", "Gender", "Education", "Occupation", 
                      "Home.Owner", "Commute.Distance", "Region", "Purchased.Bike")
bike_data[categorical_vars] <- lapply(bike_data[categorical_vars], as.factor)

# Save indices of missing values
missing_indices <- lapply(bike_data, function(x) which(is.na(x) | x == ""))

# Impute missing values using MICE
imputed_data <- mice(bike_data, m = 5, method = "pmm", seed = 123)
bike_data <- complete(imputed_data, 1)

# Check missing values after imputation
missing_bike_data_after <- sapply(bike_data, function(x) sum(is.na(x) | x == ""))
print(missing_bike_data_after)

# Display updated values
cat("Updated imputed values per variable:\n\n")
for (var in names(missing_indices)) {
  indices <- missing_indices[[var]]
  if (length(indices) > 0) {
    cat("Variable:", var, " | Count:", length(indices), "\n")
    print(data.frame(Row = indices, ImputedValue = bike_data[[var]][indices]))
  }
}
```

## 5. Checking for Outliers

```{r check-outliers, fig.width=7, fig.height=5}
# Define a function to detect outliers using the IQR method
# Define a function to detect outliers using the IQR method
detect_outliers <- function(x) {
  x_clean <- na.omit(x)
  Q1 <- quantile(x_clean, 0.25)
  Q3 <- quantile(x_clean, 0.75)
  IQR_val <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  outliers <- x_clean[x_clean < lower_bound | x_clean > upper_bound]
  return(outliers)
}

# Detect outliers and plot boxplots
outlier_vars <- c("Income", "Age", "Children", "Cars")

for (var in outlier_vars) {
  outliers <- detect_outliers(bike_data[[var]])
  cat("Outliers in", var, ":", if (length(outliers) == 0) "None" else outliers, "\n")
  cat("Count of", var, "outliers:", length(outliers), "\n\n")
  
  # Boxplot for each variable
  p <- ggplot(bike_data, aes_string(y = var)) +
    geom_boxplot(fill = "lightblue", outlier.colour = "red", outlier.shape = 16) +
    labs(title = paste("Boxplot of", var), y = var) +
    theme_minimal()
  
  print(p)
}
```

## 6. Data Visualization

```{r clean-categorical}
# Summary of Variables
str(bike_data)
summary(bike_data)
numeric_vars <- sapply(bike_data, is.numeric)
describe(bike_data[, numeric_vars])

# Pie Chart for Region
region_counts <- table(bike_data$Region)
pie(region_counts, main = "Pie Chart: Distribution of Regions", labels = paste(names(region_counts), "(", region_counts, ")"))

# Bar Chart for Education Levels
barplot(table(bike_data$Education), main = "Bar Chart: Education Levels", xlab = "Education Level", ylab = "Frequency", col = "skyblue", las = 2)

# Histogram for Income Distribution
hist(bike_data$Income, breaks = 20, main = "Histogram: Income Distribution", xlab = "Income", ylab = "Frequency", col = "red")
#scatter plot for age vs income
ggplot(bike_data, aes(x = Age, y = Income)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot: Income vs Age", x = "Age", y = "Income") +
  theme_minimal()

```

## 7. Purchased Bike Analysis

```{r purchased-bike-analysis, echo=TRUE, message=FALSE, warning=FALSE}
# Histogram of Income Variable with Summary Statistics
hist(bike_data$Income, breaks = 20, main = "Histogram: Income Distribution", xlab = "Income", ylab = "Frequency", col = "lightblue")

# Summary Statistics
income_stats <- c(Mean = mean(bike_data$Income), Median = median(bike_data$Income), Variance = var(bike_data$Income))
print(income_stats)

# Grouping Bikers by Income Ranges
bike_data$Income_Group <- cut(bike_data$Income, breaks = quantile(bike_data$Income, probs = c(0, 0.33, 0.66, 1), na.rm = TRUE),
                               include.lowest = TRUE, labels = c("Low", "Medium", "High"))
income_group_summary <- bike_data %>% 
  group_by(Income_Group) %>% 
  summarise(Total_Count = n(),
            Purchased_Count = sum(Purchased.Bike == "Yes", na.rm = TRUE),
            Purchased_Percent = round(100 * mean(Purchased.Bike == "Yes", na.rm = TRUE), 2))
print(income_group_summary)

# Correlation of Attributes with Purchased Bike
bike_data$Purchased.Bike.Num <- ifelse(bike_data$Purchased.Bike == "Yes", 1, 0)
correlations <- cor(bike_data[, sapply(bike_data, is.numeric)], use = "complete.obs")
corrplot(correlations, method = "number", title = "Correlation Matrix")

# Density Plot: Income by Bike Purchase
ggplot(bike_data, aes(x = Income, fill = Purchased.Bike)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Income by Bike Purchase", x = "Income", y = "Density")

# Clean Gender Before Plot
bike_data$Gender[bike_data$Gender == ""] <- "Missing"
bike_data$Gender <- factor(bike_data$Gender, levels = c("Male", "Female", "Missing"))

# Age vs Gender Histogram
ggplot(bike_data, aes(x = Age, fill = Gender)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  labs(title = "Age Distribution by Gender", x = "Age", y = "Count")

```

## 8. Conclusions

- **Data Cleaning:** Missing values were imputed using the MICE package. Outliers were identified using the IQR method. Categorical inconsistencies were checked and flagged.
  
- **EDA Findings:** Distributions of key variables were explored through visualizations. Income, Age, and Number of Cars showed noticeable outliers. Most purchases were made by individuals in the medium to high-income range. Mild correlations were observed between numeric features and bike purchases.


```{r final-head, echo=FALSE}
# Final look at cleaned data
head(bike_data)
```